---
title: "Evidence, privacy, and readiness controls with PhysioCompliance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evidence, privacy, and readiness controls with PhysioCompliance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PhysioCompliance supplies deterministic technical controls for physiological
records: tamper-evident audit trails, externally authenticated electronic
signatures, explicit de-identification and keyed pseudonymization, and
project-owned risk, traceability and engineering-readiness records.

These are **technical controls only**. The package does not certify that an
installation, study, organization, or submission complies with 21 CFR Part 11,
HIPAA, GDPR, IEC 62304, ISO 14971, or any other framework, and it does not
decide device status, risk acceptability, or release. Everything below runs on
synthetic data and writes only into `tempdir()`.

```{r setup}
library(PhysioCompliance)
```

## 1. A tamper-evident audit trail

Initialize a versioned SHA-256 audit chain over a record, then append events.
Verification reports tampering without modifying the object.

```{r audit}
x <- PhysioExperiment::PhysioExperiment(
  assays = list(raw = matrix(as.double(1:12), nrow = 4)),
  samplingRate = 100
)
x <- initializeAuditTrail(x, actor = "operator-01")
x <- appendAuditEvent(
  x,
  action = "quality_control",
  actor = "operator-01",
  reason = "reviewed acquisition quality"
)
verifyAuditTrail(x)
as.data.frame(auditTrail(x))[, c("sequence", "actor", "action")]
```

## 2. An externally authenticated signature

Signing and verification are delegated to caller-supplied callbacks. The package
never accepts passwords, private keys, or reusable secrets. The illustrative
callback below simply hashes the challenge; a real deployment authenticates the
signer and uses its own credential service, which determines cryptographic
strength and non-repudiation.

```{r esign}
credential <- list(id = "credential-01", algorithm = "demo-sha256",
                   fingerprint = "01:23:45:67")
sign_bytes <- function(challenge, credential) {
  digest::digest(c(challenge, serialize(credential, NULL, version = 3)),
                 algo = "sha256", serialize = FALSE, raw = TRUE)
}
x <- eSign(x, signer = "reviewer-02", meaning = "reviewed",
           credential = credential, sign = sign_bytes)

verifyESignatures(x, verify = function(challenge, signature, credential) {
  identical(signature, sign_bytes(challenge, credential))
})$valid
```

## 3. De-identification

`safeHarborPolicy()` builds a conservative field policy. The
`safe_harbor_candidate` label is a configured-field result, not a legal
determination; the applicable organization still owes the no-actual-knowledge
condition and a manual review of free text, images, and signals.

```{r deid}
rec <- PhysioExperiment::PhysioExperiment(
  assays = list(raw = matrix(as.double(1:6), nrow = 3)),
  colData = S4Vectors::DataFrame(patient_name = c("A", "B")),
  samplingRate = 100
)
clean <- deidentify(rec, safeHarborPolicy())
auditDeidentification(clean)$status
```

## 4. Keyed pseudonymization

Keys are caller-owned raw vectors of at least 32 random bytes. The returned
object carries tokens and an authenticated encrypted map, never the plaintext
map or the key.

```{r pseudo}
key <- openssl::rand_bytes(32)
p <- pseudonymize(c("subject-a", "subject-b", "subject-a"), key,
                  namespace = "study-example")
p$values
identical(reidentify(p, key), c("subject-a", "subject-b", "subject-a"))
```

Date shifting applies one deterministic per-subject offset while preserving the
exact elapsed intervals between visits:

```{r dateshift}
visits <- as.Date(c("2026-01-05", "2026-02-02", "2026-03-09"))
shifted <- dateShift(visits, subject_id = "subject-a", key = key)
all(diff(shifted) == diff(visits))
```

## 5. Project-owned risk and traceability

`riskMatrix()` records one explicit decision for every severity/probability
pair; it never multiplies ranks or supplies an acceptability threshold.

```{r riskmatrix}
severity <- data.frame(
  level_id = c("S1", "S2"), rank = 1:2,
  label = c("Minor", "Serious"),
  definition = c("Limited impact", "Significant impact"),
  stringsAsFactors = FALSE
)
probability <- data.frame(
  level_id = c("P1", "P2"), rank = 1:2,
  label = c("Unlikely", "Likely"),
  definition = c("Rarely occurs", "Often occurs"),
  stringsAsFactors = FALSE
)
decisions <- expand.grid(
  severity_id = severity$level_id, probability_id = probability$level_id,
  stringsAsFactors = FALSE
)
decisions$decision <- c("acceptable", "review_required",
                        "review_required", "unacceptable")
decisions$rationale <- paste0("POLICY-", seq_len(nrow(decisions)))
rm <- riskMatrix(severity, probability, decisions,
                 matrix_id = "RM01", version = "1.0",
                 rationale = "Project-owned decision policy")
```

`traceabilityMatrix()` normalizes a requirement-risk-control-test graph with
objective-evidence hashes, and `validateTraceability()` reports orphans,
uncontrolled risks, untested controls, and evidence-hash mismatches.

```{r trace}
evidence_root <- file.path(tempdir(), "pc-evidence")
dir.create(evidence_root, showWarnings = FALSE)
evidence_file <- file.path(evidence_root, "evidence-1.txt")
writeBin(charToRaw("objective evidence 1\n"), evidence_file)
evidence_hash <- digest::digest(evidence_file, algo = "sha256", file = TRUE)

requirements <- data.frame(
  requirement_id = "REQ1", title = "Requirement 1",
  description = "Project requirement 1", source = "project",
  source_version = "1", software_safety_class = "unclassified",
  status = "implemented", stringsAsFactors = FALSE
)
risks <- data.frame(
  risk_id = "RSK1", hazard = "Hazard 1", foreseeable_sequence = "Sequence 1",
  hazardous_situation = "Situation 1", harm = "Harm 1",
  initial_severity = "S2", initial_probability = "P2",
  initial_decision = "unacceptable",
  residual_severity = "S1", residual_probability = "P1",
  residual_decision = "acceptable",
  benefit_risk_required = FALSE, acceptance_rationale = "Project review 1",
  status = "accepted", stringsAsFactors = FALSE
)
controls <- data.frame(
  control_id = "CTL1", control_type = "inherent_safety",
  description = "Control 1", implementation_status = "verified",
  implementation_evidence = NA_character_, evidence_sha256 = NA_character_,
  stringsAsFactors = FALSE
)
tests <- data.frame(
  test_id = "TST1", level = "unit", description = "Test 1",
  expected_result = "Project criterion met", status = "pass",
  evidence_uri = basename(evidence_file), evidence_sha256 = evidence_hash,
  executed_at = "2026-07-28T03:04:01.000000Z", executor = "operator-01",
  stringsAsFactors = FALSE
)
links <- rbind(
  data.frame(source_type = "requirement", source_id = "REQ1",
             target_type = "risk", target_id = "RSK1",
             link_type = "traces_to", rationale = "Requirement-risk trace",
             stringsAsFactors = FALSE),
  data.frame(source_type = "risk", source_id = "RSK1",
             target_type = "control", target_id = "CTL1",
             link_type = "mitigates", rationale = "Risk-control trace",
             stringsAsFactors = FALSE),
  data.frame(source_type = "control", source_id = "CTL1",
             target_type = "test", target_id = "TST1",
             link_type = "verifies", rationale = "Control-test trace",
             stringsAsFactors = FALSE)
)

trace <- traceabilityMatrix(requirements, risks, controls, tests, links,
                            evidence_root = evidence_root, risk_matrix = rm)
validateTraceability(trace)$valid
```

## 6. Engineering-readiness check

`conformanceCheck()` statically parses package metadata and source. It does not
load namespaces, run tests, or launch external tools; missing execution evidence
is `NOT_EVALUATED`, never a pass. An engineering-readiness report is not a
regulatory conformity or release conclusion.

```{r conformance}
pkg <- file.path(tempdir(), "SyntheticPkg")
dir.create(file.path(pkg, "R"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(pkg, "man"), showWarnings = FALSE)
dir.create(file.path(pkg, "tests", "testthat"), recursive = TRUE,
           showWarnings = FALSE)
writeLines(c(
  "Package: SyntheticPkg", "Version: 0.1.0",
  "Title: Synthetic Example Package",
  "Description: A synthetic package used only to illustrate the readiness check.",
  "License: MIT",
  "Authors@R: person('Pat', 'Example', role = c('aut', 'cre'), email = 'pat@example.org')"
), file.path(pkg, "DESCRIPTION"))
writeLines("export(syntheticAnalysis)", file.path(pkg, "NAMESPACE"))
writeLines("syntheticAnalysis <- function(x) x",
           file.path(pkg, "R", "synthetic.R"))
writeLines(c(
  "\\name{syntheticAnalysis}", "\\alias{syntheticAnalysis}",
  "\\title{Synthetic analysis}", "\\description{Synthetic analysis.}",
  "\\usage{syntheticAnalysis(x)}",
  "\\arguments{\\item{x}{An object.}}", "\\value{An object.}",
  "\\examples{syntheticAnalysis(NULL)}"
), file.path(pkg, "man", "syntheticAnalysis.Rd"))
writeLines("test_that('synthetic', { expect_true(TRUE) })",
           file.path(pkg, "tests", "testthat", "test-synthetic.R"))

report <- conformanceCheck(pkg)
report
out <- writeConformanceReport(report, file.path(tempdir(), "conformance.json"))
basename(out)
```

## Where to go next

`?PhysioCompliance` lists every entry point. Audit-trail detection is
in-object only: to detect whole-object rollback, retain each verified
`head_hash` in a validated external append-only store and compare it on
retrieval.
